---
title: 'R-code for "No evidence of sex allocation among three sympatric plovers in Madagascar that vary in mating system"'
author: 'Maarten Vervoort, Orsolya Vincze, Callie Gesmundo, Zak Pohlen, Sama Zefania, Tamás Székely, Luke J. Eberhart-Phillips'
output:
  pdf_document:
    keep_tex: yes
  html_document:
    highlight: pygments
---
```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", cache = TRUE) # warning = FALSE
```

In this document we provide all the necessary code for reproducing the analyses presented in our paper.  To access the dataset and Rmarkdown file, please download this [GitHub](https://github.com/leberhartphillips/plover_hatching_sex_ratio) repository.  Simply follow the link and click on *Download ZIP* on the right-hand side of the page.  An explanation of the files in the repository can be found in the Readme file.  Please don't hesitate to contact Luke at `luke.eberhart[at]gmail.com` if you have any questions.

The structure of the code we present here follows the analyses presented in the *Results* section of the paper.

**Prerequisites:**

* For running the complete code you need a `files` subfolder containing the raw data downloaded from **`data`** folder provided in the [GitHub](https://github.com/leberhartphillips/plover_hatching_sex_ratio) repository.
* The following packages are needed for analysis and can be easily installed from [CRAN](http://cran.r-project.org/) by uncommenting the `install.packages` functions:

```{r, results="hide", message=FALSE, cache = TRUE}
# install.packages("dplyr")
# install.packages("lme4")
# install.packages("ggplot2")
# install.packages("RColorBrewer")
# install.packages("grid")
library(dplyr)
library(lme4)
library(ggplot2)
library(RColorBrewer)
library(grid)
```

***
## Loading and wrangling data

To start, please load the following dataset into your R environment:

- **plover chick data** contains the raw chick data collected from the field. Each row is a single capture of a chick at its nest.  Each chick has a unique identifier *chickID* and each brood has a unique identifier *nestID*. The geographic coordinates are also shown as *easting* and *northing* (Coordinate System: UTM Zone 38S Tananarive 1925) which are presented in the map to illustrate the spatial distribution of the nests used in this study. A chick can be one of three plover *species* (Kittlitz's, Madagascar, or White-fronted). *year* indicates the year during which an individual was captured and *dateinseason* indicates the number of days since the start of the breeding season that an individual hatched. The dummy variables *male* and *female* indicate the molecular sex-type of an individual with "1" signifing the sex.

```{r}
plover_chick <- 
  read.table(file = "data/plover_chick_data.txt", 
             header = TRUE, 
             stringsAsFactors = FALSE, 
             sep = "\t")
```

Define the variables of the `plover_chick` correctly: `easting` and `northing` need to recognized as numeric, and `chickID`, `nestID`, `species`, and `year` need to recognized as factors.

```{r}
plover_chick[,c("easting", "northing")] <- 
  lapply(plover_chick[,c("easting", "northing")], as.numeric)

plover_chick[,c("chickID", "nestID", "species", "year")] <- 
  lapply(plover_chick[,c("chickID", "nestID", "species", "year")], as.factor)
```

Summarize the `plover_chick` data such that each row represents data for a nest.

```{r}
plover_nest <- 
  plover_chick %>% 
  dplyr::group_by(species, year, nestID) %>%
  dplyr::summarise(no_males = sum(male),
                   no_females = sum(female),
                   Year = first(year),
                   dateinseason = first(dateinseason),
                   easting = first(easting),
                   northing = first(northing),
                   brood_size = no_males + no_females,
                   prop_male = no_males / brood_size)
```

***
## Do hatching sex ratios deviate from parity?

Linear regression mixed models for each species predicting deviation in hatching sex ratio while controlling for non-independence in siblings (i.e. random effect: `nestID`).

- **Kittlitz's plover**

```{r}
KiP_HSR_model <- lme4::glmer(cbind(no_males, no_females) ~ (1| nestID),
                          data = dplyr::filter(plover_nest, species == "Kittlitz's"),
                          family = binomial)
summary(KiP_HSR_model)
```

- **Madagascar plover**

```{r}
MP_HSR_model <- lme4::glmer(cbind(no_males, no_females) ~ (1| nestID),
                          data = dplyr::filter(plover_nest, species == "Madagascar"),
                          family = binomial)
summary(MP_HSR_model)
```

- **White-fronted plover**

```{r}
WfP_HSR_model <- lme4::glmer(cbind(no_males, no_females) ~ (1| nestID),
                          data = dplyr::filter(plover_nest, species == "White-fronted"),
                          family = binomial)
summary(WfP_HSR_model)
```

Summarize the `plover_nest` data such that each row represents a species.

```{r}
plover_summary <- 
  plover_nest %>%
  dplyr::group_by(species) %>%
  dplyr::summarise(n_chicks = sum(brood_size),
                   n_nests = length(prop_male),
                   mean = mean(prop_male),
                   SE = sd(prop_male) / sqrt(length(prop_male)),
                   lower_CI = 
                     mean(prop_male) - (sd(prop_male) / sqrt(length(prop_male)) * 1.96),
                   upper_CI = 
                     mean(prop_male) + (sd(prop_male) / sqrt(length(prop_male)) * 1.96))
```

Plot the species summary.

```{r, fig.align='center', fig.height=4, fig.width=6}
# draw the blank plot for the background
Figure_1_blank <- 
  ggplot2::ggplot(data = plover_summary, aes(y = mean, x = species)) + 
            theme_bw() +
            annotate("rect", xmin = 0, xmax = 6, ymin = 0, ymax = 0.5, alpha = 0.7,
                     fill = RColorBrewer::brewer.pal(8, "Dark2")[c(1)]) +
            annotate("rect", xmin = 0, xmax = 6, ymin = 0.5, ymax = 1, alpha = 0.7,
                     fill = RColorBrewer::brewer.pal(8, "Dark2")[c(2)]) +
                theme(text = element_text(colour = "white"),
                      legend.position = "none",
                      legend.background = element_rect(fill = NA),
                      axis.title.y = element_text(size = 12, margin = margin(0, 10, 0, 0)),
                      axis.text.y  = element_text(size = 10), 
                      axis.title.x = element_text(size = 12, margin = margin(20, 0, 0, 0)),
                      axis.text.x  = element_text(size = 10),
                      axis.ticks.x = element_blank(),
                      axis.ticks.y = element_line(size = 0.5, colour = "white"),
                      axis.ticks.length = unit(0.2, "cm"),
                      panel.grid.major = element_blank(),
                      panel.grid.minor = element_blank(),
                      panel.border = element_blank(),
                      panel.margin = unit(0.75, "lines"),
                      plot.margin = unit(c(0.5, 1.5, 0.5 ,0), "cm")) +
                scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
                scale_x_continuous(limits=c(0, 6),breaks=c(0, 1, 2), expand = c(0, 0)) +
                xlab("Plover species") +
                ylab("Hatching sex ratio (proportion male)")

# draw the plot with the data
Figure_1 <- 
  ggplot2::ggplot(data = plover_summary, aes(y = mean, x = species)) + 
            theme_bw() +
            geom_pointrange(aes(ymin = lower_CI, ymax = upper_CI)) +
            theme(legend.position = "none",
                  panel.background = element_rect(fill = "transparent",colour = NA),
                  plot.background = element_rect(fill = "transparent",colour = NA),
                  axis.title.y = element_text(size = 12, margin = margin(0, 10, 0, 0)),
                  axis.text.y  = element_text(size = 10), 
                  axis.title.x = element_text(size = 12, margin = margin(20, 0, 0, 0)),
                  axis.text.x  = element_text(size = 10),
                  axis.ticks.x = element_blank(),
                  axis.ticks.y = element_line(size = 0.5, colour = "grey40"),
                  axis.ticks.length = unit(0.2, "cm"),
                  panel.grid.major = element_blank(),
                  panel.grid.minor = element_blank(),
                  panel.border = element_blank(),
                  panel.margin = unit(0.75, "lines"),
                  plot.margin = unit(c(0.5, 1.5, 0.5, 0), "cm")) +
            scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
            xlab("Plover species") +
            ylab("Hatching sex ratio (proportion male)")

# Print the two layers
grid::pushViewport(
  grid::viewport(
    layout = grid::grid.layout(1, 1, widths = unit(1, "npc"))))
print(Figure_1_blank, newpage = FALSE)
print(Figure_1, newpage = FALSE)
grid::popViewport()
```

***
## Does hatching sex ratio vary seasonaly or annually?

Linear regression mixed models for each species predicting annual and seasonal deviations in hatching sex ratio while controlling for non-independence in siblings (i.e. random effect: `nestID`).

- **Kittlitz's plover**

```{r}
# Annual model
KiP_HSR_year <- 
  lme4::glmer(male ~ (1| nestID) + year, 
              data = dplyr::filter(plover_chick, species == "Kittlitz's"),
              family = binomial)

# Seasonal model
KiP_HSR_season <- 
  lme4::glmer(male ~ (1| nestID) + dateinseason, 
              data = dplyr::filter(plover_chick, species == "Kittlitz's"),
              family = binomial)

# No temporal effect model
KiP_HSR_null <- 
  lme4::glmer(male ~ (1| nestID), 
              data = dplyr::filter(plover_chick, species == "Kittlitz's"),
              family = binomial)

# Assess the effect of year
anova(KiP_HSR_year, KiP_HSR_null)

# Assess the effect of season
anova(KiP_HSR_season, KiP_HSR_null)
```

- **Madagascar plover**

```{r, warning=FALSE}
# Annual model
MP_HSR_year <- 
  lme4::glmer(male ~ (1| nestID) + year, 
              data = dplyr::filter(plover_chick, species == "Madagascar"),
              family = binomial)

# Seasonal model
MP_HSR_season <- 
  lme4::glmer(male ~ (1| nestID) + dateinseason, 
              data = dplyr::filter(plover_chick, species == "Madagascar"),
              family = binomial)

# No temporal effect model
MP_HSR_null <- 
  lme4::glmer(male ~ (1| nestID), 
              data = dplyr::filter(plover_chick, species == "Madagascar"),
              family = binomial)

# Assess the effect of year
anova(MP_HSR_year, MP_HSR_null)

# Assess the effect of season
anova(MP_HSR_season, MP_HSR_null)
```

- **Madagascar plover**

```{r, warning=FALSE}
# Annual model
WfP_HSR_year <- 
  lme4::glmer(male ~ (1| nestID) + year, 
              data = dplyr::filter(plover_chick, species == "White-fronted"),
              family = binomial)

# Seasonal model
WfP_HSR_season <- 
  lme4::glmer(male ~ (1| nestID) + dateinseason, 
              data = dplyr::filter(plover_chick, species == "White-fronted"),
              family = binomial)

# No temporal effect model
WfP_HSR_null <- 
  lme4::glmer(male ~ (1| nestID), 
              data = dplyr::filter(plover_chick, species == "White-fronted"),
              family = binomial)

# Assess the effect of year
anova(WfP_HSR_year, WfP_HSR_null)

# Assess the effect of season
anova(WfP_HSR_season, WfP_HSR_null)
```